---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# Nchalo Water Project – Household Survey 2021

<!-- badges: start -->

[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)

[![DOI](https://zenodo.org/badge/DOI/.svg)](https://zenodo.org/doi/)
<!-- badges: end -->

This dataset was collected in 2021 using the mWater mobile data
collection platform, as part of the Nchalo Water Project. The survey
captures household-level and community-level information related to
water access, livelihoods, governance, infrastructure, and service
delivery in rural communities around Nchalo, Malawi.

The data reflects conditions at the time of collection and is intended
to support targeted decision-making around water supply projects and
integrated rural development efforts.

1.  **Use cases of the data** \
    Local Government Authorities (District Water Offices, Planning
    Departments) To inform localized infrastructure development and
    service delivery planning.

2.  Non-Governmental Organizations (NGOs) To support rural WASH,
    livelihood improvement, and community empowerment programs.

3.  Donor Agencies and Development Partners For targeting investments
    and evaluating the impact of funded projects.

4.  Researchers and Academics For studying rural development, community
    resilience, water governance, and socio-economic trends.

5.  Community-Based Organizations (CBOs) To advocate for resources, plan
    community-led initiatives, and participate in decision-making.

6.  Monitoring & Evaluation (M&E) Specialists To use as baseline data
    for future assessments or project comparisons.

## Installation

You can install the development version of nchalohhwatersurvey from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/nchalohhwatersurvey")
```

```{r}
## Run the following code in console if you don't have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

1.  Click Download CSV. A window opens that displays the CSV in your
    browser.
2.  Right-click anywhere inside the window and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/nchalohhwatersurvey/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to community-level household survey
conducted as part of the Nchalo Water Project. Its purpose is to assess
livelihoods, water usage, community governance, infrastructure access,
and opportunities for value addition in the context of a rural water
supply or WASH-related development intervention.

```{r}
library(nchalohhwatersurvey)
```

### nchalohhwatersurvey

The dataset `nchalohhwatersurvey` contains `r nrow(nchalohhwatersurvey)`
observations and `r ncol(nchalohhwatersurvey)` variables

```{r}
nchalohhwatersurvey |> 
  head(3) |> 
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "nchalohhwatersurvey.rda") |>
  dplyr::select(variable_name:description) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```

## Data Visualization Examples

```{r}
library(nchalohhwatersurvey)
# required Libraries
library(tidyverse)
library(maps)
library(ggplot2)
library(dplyr)

# Vizualisation 1: Main sources of drinking water
# Count the frequency of each main drinking water source in the dataset
water_source_counts <- nchalohhwatersurvey %>%
  # Remove rows where main_drinking_water_source is missing
  filter(!is.na(main_drinking_water_source)) %>%
  # Count occurrences of each unique main drinking water source
  count(main_drinking_water_source) %>%
  # Arrange counts in descending order for better visualization
  arrange(desc(n))

# Create a bar chart of the water source counts
ggplot(water_source_counts, aes(x = reorder(main_drinking_water_source, -n), y = n)) +
  # Use geom_bar with identity stat to plot counts
  geom_bar(stat = "identity", fill = "#2c7fb8") +
  # Add chart title and axis labels
  labs(
    title = "Main Sources of Drinking Water",
    x = "Drinking Water Source",
    y = "Number of Communities"
  ) +
  # Use a minimal theme for a clean look
  theme_minimal() +
  # Rotate x-axis labels 45 degrees for readability
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Vizualisation 2: Map with points colored by main drinking water source
# Load Malawi map data from maps package
malawi_map <- map_data("world", region = "Malawi")

# Filter your data for plotting (remove missing coords or water sources)
plot_data <- nchalohhwatersurvey %>%
  filter(!is.na(latitude), !is.na(longitude), !is.na(main_drinking_water_source))

# Plot Malawi map with points colored by water source
ggplot() +
  geom_polygon(data = malawi_map, aes(x = long, y = lat, group = group),
               fill = "gray95", color = "black") +
  geom_point(data = plot_data, 
             aes(x = longitude, y = latitude, color = main_drinking_water_source),
             alpha = 0.7, size = 3) +
  scale_color_brewer(palette = "Set2") +
  coord_fixed(1.3) +  # fix aspect ratio
  labs(
    title = "Main Drinking Water Sources in Malawi Communities",
    x = "Longitude",
    y = "Latitude",
    color = "Water Source"
  ) +
  theme_minimal()

```

## License

Data are available as
[CC-BY](https://github.com/openwashdata/%7B%7B%7Bpackagename%7D%7D%7D/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("nchalohhwatersurvey")
```
